Data Management for LSST Special Programs

### 2024-03-01 DMTN-065.tex

\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~P.~Guy, and the Data Management System Science Team}

>>>>

\author{M.~L.~Graham, Y.~AlSayyad, K.~Bechtol, E.~C.~Bellm, J.~F.~Bosch, J.~L.~Carlin, 
G.~P.~Dubois-Felsmann, L.~P.~Guy, M.~Juri\'{c}, K.-T.~Lim, E.~S.~Rykoff, C.~T.~Slater,
and the Data Management System Science Team}

### 2024-03-01 ap_spcs.tex

\textbf{\emph{Hypothetical}} examples of Special Programs
and the regular, Special, and User-Generated processing
to illustrate what \emph{might be done}.

@@

The steps used to describe the hypothetical processing for each case scenario are: \\
Step 1. Data acquisition. \\
Step 2. Regular Prompt processing and Alert Production. \\
Step 3. Special Processing with reconfigured pipelines. \\
Step 4. Regular processing for inclusion in Main Survey data products. \\
Step 5. User-Generated Processing. \\

@@

Step 2. Regular Prompt processing and Alert Production. \\
Each $2\times15$ second standard visit is processed by the Prompt pipeline 
and alerts are released within 60 seconds.
Within 24 hours, the {\tt DiaSource} and {\tt DiaObject} catalogs are updated
to include the results of Prompt processing of these visits.
After 80 hours, the processed visit images and difference images become available.
All images and sources originating from this Special Program have 
region and observing mode labels, e.g., ``SP-OSSO".

@@

Step 4. Regular processing for inclusion in Main Survey data products. \\
Every year, each $2\times15$ second standard visit is reprocessed by the DIA data release pipelines
and the results are included alongside Main Survey data in the relevant DIA data products
(e.g., processed visit images, difference images, associated source catalogs).
In the first year after the Special Program is executed,
Rubin Data Management finds that 10\% of the standard visits from this Special Program
had coordinates and image quality that help improve uniformity of the all-sky coadd,
and so they are included.
In later years, this fraction decreases (remember, this is \emph{hypothetical}).
In all data releases, any and all processed images and catalog sources that originate in visits from this Special Program
have the same region and observing mode labels, e.g., ``SP-OSSO".

@@

Step 2. Regular Prompt processing and Alert Production. \\
Same as above, but the observing mode label might be, e.g., ``SP-DDF-COSMOS".

@@

Step 4. Regular processing for inclusion in Main Survey data products. \\
Every year, each standard visit is reprocessed by the DIA data release pipelines
and the results are included alongside Main Survey data in the relevant DIA data products
(e.g., processed visit images, difference images, associated source catalogs).
Due to their small dither and lack of rotation, not even a single DDF image 
is used to supplement the Main Survey's all-sky coadd.

@@

Step 2. Regular Prompt processing and Alert Production. \\
Pending studies of DIA and Alert Production pipeline capabilities to process 
short-exposure, high sky-background images (see Section~\ref{ssec:proc_bounds_processing}).

@@

Step 4. Regular processing for inclusion in Main Survey data products. \\
These short-exposure, high sky background images would not contribute to the data products created for the Main Survey.

>>>>

\textbf{\emph{Hypothetical}} examples of Special Programs
and the standard, Special, and User-Generated processing
to illustrate what \emph{might be done}.

@@

The steps used to describe the hypothetical processing for each case scenario are: \\
Step 1. Data acquisition. \\
Step 2. Standard Prompt processing and Alert Production. \\
Step 3. Special Processing with reconfigured pipelines. \\
Step 4. Standard processing for inclusion in Main Survey data products. \\
Step 5. User-Generated Processing. \\

@@

Step 2. Standard Prompt processing and Alert Production. \\
Each $2\times15$ second standard visit is processed by the Prompt pipeline 
and alerts are released within 60 seconds.
Within 24 hours, the {\tt DiaSource} and {\tt DiaObject} catalogs are updated
to include the results of Prompt processing of these visits.
After 80 hours, the processed visit images and difference images become available.
All images and sources originating from this Special Program have 
region and observing mode labels, e.g., ``SP-OSSO".

@@

Step 4. Standard processing for inclusion in Main Survey data products. \\
Every year, each $2\times15$ second standard visit is reprocessed by the DIA data release pipelines
and the results are included alongside Main Survey data in the relevant DIA data products
(e.g., processed visit images, difference images, associated source catalogs).
In the first year after the Special Program is executed,
Rubin Data Management finds that 10\% of the standard visits from this Special Program
had coordinates and image quality that help improve uniformity of the all-sky coadd,
and so they are included.
In later years, this fraction decreases (remember, this is \emph{hypothetical}).
In all data releases, any and all processed images and catalog sources that originate in visits from this Special Program
have the same region and observing mode labels, e.g., ``SP-OSSO".

@@

Step 2. Standard Prompt processing and Alert Production. \\
Same as above, but the observing mode label might be, e.g., ``SP-DDF-COSMOS".

@@

Step 4. Standard processing for inclusion in Main Survey data products. \\
Every year, each standard visit is reprocessed by the DIA data release pipelines
and the results are included alongside Main Survey data in the relevant DIA data products
(e.g., processed visit images, difference images, associated source catalogs).
Due to their small dither and lack of rotation, not even a single DDF image 
is used to supplement the Main Survey's all-sky coadd.

@@

Step 2. Standard Prompt processing and Alert Production. \\
Pending studies of DIA and Alert Production pipeline capabilities to process 
short-exposure, high sky-background images (see Section~\ref{ssec:proc_bounds_processing}).

@@

Step 4. Standard processing for inclusion in Main Survey data products. \\
These short-exposure, high sky background images would not contribute to the data products created for the Main Survey.

### 2024-03-01 sec_proc.tex

Appendix~\ref{sec:spcs} provides detailed examples for the processing
of data from Special Programs, including scenarios in which regular,
Special, and User-Generated Processing are all involved.

@@

\subsection{Regular (non-special) processing}\label{ssec:proc_reg}

Recall that the term ``regular processing" is used only within this document,
and refers to the image processing pipelines that are designed for, 
and will be applied to, the Main Survey's observations (Section~\ref{sec:intro}).

Decisions about when to apply regular processing to Special Programs data,
or when to include it in the data products for the Main Survey
(e.g., if it improves the all-sky coadds), 
are ultimately left to the discretion of the Rubin Observatory's 
Data Management and System Performance departments in Rubin Operations.

@@

\textbf{How Main Survey and Special Program data would co-exist.}
No ``unique and separate" data products for the Special Progams would be 
produced by regular Prompt processing.
Special Programs data that is processed by the Prompt pipeline would 
contribute to the Prompt data products for the Main Survey as 
described in Section 3 of the DPDD \citedsp{LSE-163}. 
These data products are the results of Difference Image Analysis (DIA),
such as the difference images, catalogs of sources detected in difference
images ({\tt DiaSources}) and associated static-sky {\tt DiaSources}
into {\tt DiaObjects}, and alert packets.
Including visits from Special Programs in regular Prompt processing alongside
vists from the Main Survey is not, in general, anticipated to affect Main Survey 
or WFD-specific science goals.
For example, analyses for a WFD-only subset could still be done using the observing mode
and region labels described in Section~\ref{ssec:sci_labels}, which would be
propogated to difference images, difference-image catalogss, and alerts.

@@

\item \textbf{Possible but not necessary:}
a time-domain mini-survey that uses standard visits \emph{could}
have separate difference-image analysis object and source catalogs
generated, but this is not necessary as the science goals for the
mini-survey can be met by processing its data with regular Prompt
Processing, and ensuring the data is properly labeled.

>>>>

Appendix~\ref{sec:spcs} provides detailed examples for the processing
of data from Special Programs, including scenarios in which standard,
Special, and User-Generated Processing are all involved.

@@

\subsection{Standard (non-special) processing}\label{ssec:proc_reg}

Recall that the term ``standard processing" is used only within this document,
and refers to the image processing pipelines that are designed for, 
and will be applied to, the Main Survey's observations (Section~\ref{sec:intro}).

Decisions about when to apply standard processing to Special Programs data,
or when to include it in the data products for the Main Survey
(e.g., if it improves the all-sky coadds), 
are ultimately left to the discretion of the Rubin Observatory's 
Data Management and System Performance departments in Rubin Operations.

@@

\textbf{How Main Survey and Special Program data would co-exist.}
No ``unique and separate" data products for the Special Progams would be 
produced by standard Prompt processing.
Special Programs data that is processed by the Prompt pipeline would 
contribute to the Prompt data products for the Main Survey as 
described in Section 3 of the DPDD \citedsp{LSE-163}. 
These data products are the results of Difference Image Analysis (DIA),
such as the difference images, catalogs of sources detected in difference
images ({\tt DiaSources}) and associated static-sky {\tt DiaSources}
into {\tt DiaObjects}, and alert packets.
Including visits from Special Programs in standard Prompt processing alongside
vists from the Main Survey is not, in general, anticipated to affect Main Survey 
or WFD-specific science goals.
For example, analyses for a WFD-only subset could still be done using the observing mode
and region labels described in Section~\ref{ssec:sci_labels}, which would be
propogated to difference images, difference-image catalogss, and alerts.

@@

\item \textbf{Possible but not necessary:}
a time-domain mini-survey that uses standard visits \emph{could}
have separate difference-image analysis object and source catalogs
generated, but this is not necessary as the science goals for the
mini-survey can be met by processing its data with standard Prompt
Processing, and ensuring the data is properly labeled.

### 2024-03-01 sec_sci.tex

\textbf{All visits should have a region label \textit{and} an observing mode 
label, and these labels should be propagated to processed images and catalog data.}

\textbf{Region Labels -- } 
Based on Right Acension and Declination.
To be applied to data by Rubin Data Management, e.g., using a look-up table.
Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
Region labels should be propagated to processed images (visit, 
difference, and deep coadds), catalog sources, and catalog static-sky objects.

\textbf{Observing Mode Labels -- } 
Based on the scheduler mode at the time of the observation
(i.e., survey cadence or strategy).
To be applied by the Survey Scheduler team via the observations scheduler.
Observing mode labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
options like engineering, commissioning, and director's discretionary.
Observing mode  labels should be propagated to single-visit processed and 
difference images and catalog sources.
It would not make sense to propagate observing mode labels to deep coadds or 
object tables, as these data products could be a mix of modes.

@@

\item To enable users to query and retrieve processed image and catalog
data associated with a specific Special Program, and meet the science goals
of that Special Program, when regular processing as been applied 
(e.g., Prompt Processing, Section~\ref{ssec:sci_pproc}).

\item To enable provenance when Special Programs data in included in regular
processing and the Main Survey's data products (e.g., if used to improve the
all-sky coadd).

@@

\textbf{Necessary -- } When the primary science goal for a Special Program 
cannot be met by including the data in regular processing (e.g., Prompt processing),
or where doing so would compromise the Main Survey data products (e.g., introduce additional non-uniformity beyond what would be expected based on the survey strategy).

@@

Further discussion of regular Prompt processing of Special Programs data, with
examples and a discussion of a few challenges,
is provided in Section~\ref{sssec:proc_reg_prompt}.

>>>>

\textbf{All visits should have at least one region label \textit{and} one observing mode 
label, and these labels should be propagated to processed images and catalog data.}

\textbf{Region Labels -- } 
Based on Right Acension and Declination.
To be applied to data by Rubin Data Management, e.g., using a look-up table.
Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
Region labels should be propagated to processed images (visit, 
difference, and deep coadds), catalog sources, and catalog static-sky objects.
A given image might have more than one region label due to overlaping regions,
but they could be, e.g., held in a single string as a CSV.

\textbf{Observing Mode Labels -- } 
Based on the scheduler mode at the time of the observation
(i.e., survey cadence or strategy).
To be applied by the Survey Scheduler team via the observations scheduler.
Observing mode labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
options like engineering, commissioning, and director's discretionary.
Observing mode labels should be propagated to single-visit processed and 
difference images and catalog sources.
It would not make sense to propagate observing mode labels to deep coadds or 
object tables, as these data products could be a mix of modes.
It is left to the discretion of the Survey Scheduler team whether more than one
observing mode labels might be applied at a time and, e.g., held in a single string as a CSV.

A given image would probably only have one observing mode label

@@

\item To enable users to query and retrieve processed image and catalog
data associated with a specific Special Program, and meet the science goals
of that Special Program, when standard processing as been applied 
(e.g., Prompt Processing, Section~\ref{ssec:sci_pproc}).

\item To enable provenance when Special Programs data in included in standard
processing and the Main Survey's data products (e.g., if used to improve the
all-sky coadd).

@@

\textbf{Necessary -- } When the primary science goal for a Special Program 
cannot be met by including the data in standard processing (e.g., Prompt processing),
or where doing so would compromise the Main Survey data products (e.g., introduce additional non-uniformity beyond what would be expected based on the survey strategy).

@@

Further discussion of standard Prompt processing of Special Programs data, with
examples and a discussion of a few challenges,
is provided in Section~\ref{sssec:proc_reg_prompt}.
